Policing for Profit and the Undermining of Democracy

### 2021-04-19 markdown/draft_paper.Rmd
write
lots of changes

---
title: "Policing for Profit and the Mobilization of Black Voters"
subtitle: "Fines, Fees, and Turnout"
author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
- Kevin Morris^[PhD Student, CUNY Graduate Center, Department of Sociology (kmorris@gradcenter.cuny.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  The American criminal legal system is an important site of political socialization: sociologists show how the alienating effect of criminalization manifests as systems avoidance, and political scientists have measured the effects of policing and incarceration on political participation. Despite this burgeoning literature, no research has directly investigated how ticketing for low-level offenses affects political participation. Using data from the Census of Governments and the national voter file, we test the causal effect of fines and fees practices on American voting behavior at the city level. Additionally, we exploit individual-level ticketing data from Hillsborough County, Florida between 2003 and 2019 to determine whether getting a ticket impacts political participation. Nationally, we estimate that each 10 percent increase in city fines and fees collections between 2012 and 2017 is associated with a roughly 0.6 percentage point decrease in non-Black turnout but was not associated with Black turnout. In Tampa, Florida, we estimate that being ticketed increases a non-Black individual’s likelihood of voting by 3.7 percentage points, and that the effect is larger for Black voters.

  |
  |
  **Keywords**: voting, monetary sanctions, race, punishment, political participation
nocite: | 
  @Burch2011, @Burch2013, @White2019a

@@

Work exploring how criminalization directly and indirectly influences political participation has exploded in recent years. This work, however, has largely focused on the effects of highly consequential contact with the criminal legal system such as incarceration and felony convictions [@Burch2014; @Lee2014]. Although some work has leveraged survey data and qualitative approaches to interrogate lower-level police interactions---such as ticketing---our project represents the first use of individual-level administrative data to identify the causal effect of ticketing on voter behavior. We begin by joining fees and fines data from municipalities around the country in 2012 and 2017 with a novel dataset estimating racial turnout for each municipality in the country. While we find that increased fees and fines likely cause lower turnout among non-Black Americans, we find no effect of fees and fines on Black turnout.

After exploring national data, we use individual-level traffic stop data from Hillsborough County, Florida, to identify the turnout patterns of voters who were ticketed between the 2012 and 2018 elections. By matching individuals who were ticketed to other, similar voters who were not ticketed and running a difference-in-differences model, we estimate the causal effect of these tickets on turnout. Using self-reported race from the Florida registered voter file, we identify large, *positive* treatment effects for all voters---effects that were especially large for Black voters..

We demonstrate that police ticketing practices---the most widespread form of police contact in America---influence the political behavior of Americans, with distinct effects for Black and non-Black voters. This finding supports and extends Hannah Walker’s theory of criminal legal contact is mobilizing / less demobilizing for nonwhite political participation [@Walker2020; @Walker2014]. Our results also complicate Vesla Weaver and Amy Lerman’s theory of low-level criminal legal contact discouraging political participation [@Weaver2010]. Our findings are relevant for interdisciplinary scholars of crime, race, politics, political economy, social movements, and policing.

@@

Sociologists have incorporated contemporary fines and fees practices into a broader empirical-historical theory of racialized punishment policy [@Harris2010; @Friedman2020]. Additionally, an emerging political science literature directly investigates the relationship between criminalization and political behavior.

@@

Another process by which criminalization shapes the electorate is “prison gerrymandering,” or the practice of counting incarcerated people as residing in the district where their prison is located for the purposes of creating electoral districts. This practice inflates the voting power of rural and white voters while diminishing the power of urban, Black, and Latino voters [@Remster2018].

@@

Using existing national survey data and extended interviews, Vesla Weaver and Amy Lerman found that trust in government and willingness to vote decrease as individuals progress through increasingly intense levels of criminal legal contact (questioned by police, arrested, convicted, incarcerated) [@Weaver2010]. Weaver and Lerman also investigate the potential role of criminal legal contact in facilitating a collective political consciousness among Black Americans, or the concept of a “linked fate” where “one’s individual opportunities are linked to the entire racial group” [@Lerman2014, page 267]. Their interview data lead them to conclude that while criminal legal contact does increase perceptions of linked fate among Black respondents, “a custodial linked fate is not the building block of greater activism and a mobilized racial identity; instead, it is demobilizing”---and interview respondents who had been incarcerated reported that withdrawal from public institutions and democratic participation constituted a form of self-preservation [@Lerman2014].

@@

It may be the case that Black and non-Black Americans employ different self-preservation strategies, however. Ariel White’s analysis of how short jail sentences affect voting in Harris County, Texas found that Black defendant turnout in subsequent elections dropped by 13 percentage points---but the effect on white turnout was not statistically significant [@White2019].

@@

Importantly, Walker found that nonwhite people who had experienced *either* proximal or personal contact *at any level of intensity*, including being stopped and questioned by police, were mobilized to take part in nonvoting political activity---and the same held true for whites, but with a much smaller effect size. Walker theorizes that this finding suggests that nonwhite people view criminalization “as a form of systemic political threat,” an analysis comparable to “linked fate.” Notably, Walker’s survey questions excluded “minor traffic stops,” the primary focus of this paper.

@@

Although geographic or spatial analyses of how criminalization affects political socialization technically move beyond proximal contact as an extant conceptual framework, we argue that geographic proximity should work similarly to social network proximity---if one neighborhood is heavily policed, people who live in that neighborhood are more likely to have social ties who have experienced policing. Traci Burch found that North Carolina block groups with higher concentrations of people who had experienced incarceration or criminal legal supervision saw substantial reductions in voter turnout, and confirmed her findings at the individual level [@Burch2014]. @Morris2020 similarly finds that neighborhoods with more active voters lost to felony disenfranchisement turned out at lower rates in the 2017 mayoral race in NYC, and that this effect was particularly pronounced in Black neighborhoods.

@@

Over the past few years, scientific understanding of the political effects of more serious interactions with the criminal legal system has grown by leaps and bounds. Table \ref{tab:lit-table} summarizes our understanding of the literature. As Table \ref{tab:lit-table} makes clear, a large body of research has now demonstrated the negative effects of high-level contact with the criminal legal system on voter turnout. Although lower-level stops are far more frequent, our knowledge of their (de)mobilizing effects is under-developed.

@@

Apart from the fact that no prior work has directly investigated the relationship between police ticketing practices and political behavior, we argue that fines and fees invoke a distinct historical legacy of racist exploitation that merits further study.

@@

Racial capitalist political economy imposes fiscal constraints regardless of city officials’ intentions. In their study of fines and fees reliance in Chicago suburbs, Josh Pacewicz and John Robinson, III show how the racialization of municipal opportunity structures prevents majority-Black suburban governments from acquiring sufficient revenue from “good” sources such as sales taxes or property taxes, catalyzing a reliance on fines and fees to fund essential government services [@Pacewicz2020]. Importantly, Pacewicz and Robinson III note that these racialized municipal opportunity structures are tied to the political-historical processes of neoliberalism and fiscal austerity, an observation that aligns with recent scholarship detailing the origins of fees paid by incarcerated people [@Kirk2020]. The late 20th century “property tax revolt” most significantly benefited white homeowners [@Martin2017], and without property tax hikes as a revenue-raising option, smaller cities and towns have few choices.

@@

With this history in mind, we argue that Black Americans can situate exploitative ticketing practices in a longer historical legacy of collective disparate treatment and violence at the hands of the state, which may facilitate a stronger collective political analysis when police ticketing practices become more extreme. @Walker2020 makes clear that, when Black Americans locate their contact with the criminal legal system in a narrative of racial injustice, they can be mobilized to vote. The historical use of fees and fines make this sort of police contact especially likely to evoke such narratives. Accordingly, we expect that Black voters’ political behavior in response to police ticketing will be distinct from the political behavior of white voters.

@@

**Census of Governments**: Every five years, the US Census Bureau collects data from governmental bodies around the country. We use two snapshots of the COG: one that covers the years 2008 - 2012, and another that covers 2013 - 2017. The COG includes a host of information about state and municipal finances. The COG asks municipalities to report how much revenue they collect from “penalties imposed for violation of law; civil penalties (e.g., for violating court orders); court fees if levied upon conviction of a crime or violation... and forfeits of deposits held for performance guarantees or against loss or damage (such as forfeited bail and collateral).” This data has been used as a proxy for the amount of tickets issued at the local level in scholarly research [see, for instance, @Goldstein2020; @Sances2017].

The COG data are useful to isolate ticketing practices because other fines and fees (e.g. fines imposed at sentencing in state court, probation fees, conviction fees, public defender fees) are typically collected by county or state governments. The COG data only include local revenues, which means the share of fines revenue corresponds more directly to local police practices. Following the guidance of @Sances2017, we retain only municipalities with police and/or court systems, and only municipalities with populations of at least 2,500.

**Registered Voter Files**: The COG data do not include estimates of turnout at the local level. To estimate local turnout, we leverage registered voter files provided by data vendor L2. Voters in Washington, DC, and every state in the country (with the exception of North Dakota) requires voters to register to vote prior to participating. These files are considered public record, and in most states contain information about each voter’s age, residential address, political affiliation, and participation history. L2 collects these voter files from around the country and merges them with proprietary data that estimates other voter- and household-level characteristics such as race and education. L2 also geocodes voters’ residential addresses. L2 does not collect all states’ voter file on the same date; the date of the “raw” voter file from each state from which the L2 data are modelled can be found in Appendix A.

We merge these geocoded voter files with “place” and “county subdivision” shapefiles published by the US Census Bureau. By aggregating these individual-level records to the municipal level, we are able to estimate the number of registered voters and the number of ballots cast in every city in the country. By using the racial estimates provided by L2, we can estimate the number of ballots cast by voters of each race in each city. Although each voters’ race is modeled based on L2’s algorithms---and is not self-reported---it is possibly incorrect for some voters. We assume this does not present a problem when records are aggregated to the municipal level.

Although turnout could be calculated as a share of *registered voters* who participated, such an approach could be biased. If fees and fines collection impact the number of individuals who register to vote, but not the share of registered voters who participate, our estimates will be compromised. We therefore calculate turnout by dividing the number of ballots cast according to L2 by the 5-year estimate of each municipality’s citizen voting-age population (CVAP) from the US Census Bureau. These estimates are available for individual racial groups. Because our turnout numerator and denominators come from different sources there is slight measurement error: our estimate indicates, for instance, that roughly 1 percent of municipalities had Black turnout above 100 percent. Where more votes were cast than the Census estimate for CVAP, we recode turnout as 100 percent. Our results do not change if we instead simply remove these municipalities.

**Estimation Strategy**: We begin by merging each municipality’s fees and fines per capita from the 2012 COG to its turnout from the 2014 midterm election, and the 2017 COG to its 2018 turnout. Because this measure is highly right-skewed, we use the natural log of fees and fines per capita plus one. We also control for other municipal and state characteristics. These include racial composition, population density, median income, collegiate education, median age, and share of the population over 64 years of age from the 5-year American Community Survey ending in 2014 and 2018, and total municipal revenue, share of revenue that comes from taxes, and share of each municipality’s revenue that comes from the state or federal government in the 2012 and 2017 COGs. We retain only municipalities whose data is reported to the COG in both 2012 and 2017.

We then run a two-way fixed effects model to explore whether changes in fees and fines collections from 2012 to 2017 were associated with changes in turnout, and whether these effects differed for Black and non-Black voters.

Unfortunately, the 5-year cycle of the COG does not neatly line up with the general 4-year cycle typical of American elections. The timing of the COG and midterm elections mean that our analytical strategy is conservative. If municipalities shift their fees and fines collections dramatically over the 2013 -- 2017 period, this shift may influence their 2014 turnout. However, we assume that turnout in 2014 will still be influenced by the policing practices of the 2008 -- 2012 period, while 2018 turnout will be largely influenced by the practices of the 2013 -- 2017 period.

@@

*Hypothesis 1: Increased reliance on fines and fees revenue at the city level affects voter turnout at the city level.*

*Hypothesis 2: The effect of the city-level relationship between fines and voter turnout is less demobilizing or more mobilizing for Black voters, given the historical and explicit racial injustice narratives used to explain these practices.*

@@

Our national test of the relationship between municipal fees and fines collection and turnout offers novel insight into the political socialization of low-level police contact. It cannot, however, directly observe the causal mechanism we argue is at play: namely, that direct contact with the criminal legal system influences turnout. Our national analysis suffers from what is called the “ecological fallacy:” it is possible that any observed relationship between fees and fines collections and municipal turnout is being driven by the behavior of voters other than those who are ticketed.

To test whether a ticket directly influences the turnout voters we turn to individual-level data from Hillsborough County, Florida (home to Tampa). Hillsborough County is interesting for theoretical reasons: Tampa is in the top decile of municipalities in terms of fees and fines increase between the 2012 and 2017 COGs, and its population puts it in the top 1 percent of cities nationwide. It also has a relatively large Black population, allowing us to test for racial differences in the turnout effect of ticketing.

@@

Hillsborough County also makes a unique set of data publicly available. The county clerk makes information available about every traffic stop in the county going back to 2003. This data includes the name and date of birth of the individual stopped; the date of the offense; and the amount the individual paid for a ticket. We retain only the records where the individual paid money.

To test the effect of ticketing on individual-level turnout, we match the first and last names and dates of birth against the Hillsborough County registered voter file. Here, rather than rely on a snapshot from L2, we use the raw voter file directly from the state. Using a single post-treatment snapshot of the voter file can result in conditioning on a post-treatment status [see @Nyhan2017]. Instead, we collect snapshots of the voter file following each even-year general election between 2012 and 2018. We thus observe virtually all individuals who were registered  to vote at any time during our period of study. Unique voter identification numbers allow us to avoid double-counting voters who are registered in multiple snapshots. We retain each voter’s earliest record, and geocode voters to their home census block groups.

@@

By matching the police stop and voter records, we identify all voters who were ticketed between the 2012 and 2018 general elections. Voters are considered “treated” for every general election following their first ticket. This means that a voter who was ticketed in early 2014 was considered treated for the 2014, 2016, and 2018 elections. A voter treated in 2017, on the other hand, is only considered treated in the 2018 election.

These voters are then matched to two control voters using what methodologists call a “genetic” process [@Sekhon2011]. The pool of potential control voters are the individuals who are never stopped during the 2012 -- 2018 period, regardless of whether they paid a fine. We match voters on individual-level characteristics (race / ethnicity; gender; party affiliation; age; registration date; number of traffic stops between 2003 and the 2012 election) and block group-level characteristics from the 2012 5-year ACS estimates (median income; share of the population with some college; unemployment rate).

We also match treated and control voters on their 2010 and 2012 general election turnout, as well as any other even-yeared general election turnout in the pre-treatment period. In other words, a voter ticketed in 2013 is matched to controls based on turnout in 2010 and 2012, while a voter ticketed in 2017 is matched to their control using 2010, 2012, 2014, and 2016 turnout. Because voters treated in later periods are matched to their controls along more covariates, we run three separate matches: one for voters stopped between the 2012 and 2014 elections; for voters stopped between the 2014 and 2016 elections; and for those stopped between the 2016 and 2018 elections.

We then run ordinary least squares difference-in-difference regressions to test the following hypotheses.

*Hypothesis 3: Among individuals who were ticketed (“treated”) in Tampa, their likelihood of turning out in a subsequent election roughly matches the racioethnic distribution apparent in the national city-level data. If fines uniquely mobilize Black voters nationally, they should also uniquely mobilize Black voters in Tampa.*

<!-- Because Tampa holds off-year local elections, we are able to test whether mobilized voters target officials who are able to change local police practices. @Laniyonu2019 gives us reason to believe that the effect should be stronger for local elections.  -->

<!-- *Hypothesis 4: Any potential mobilizing turnout effect in Tampa is stronger for local (mayoral) elections, since those elections directly affect police practices.* -->

@@

Figure \ref{fig:did1} makes a number of things immediately apparent. First, we can see that for both Black and non-Black voters the parallel trends assumption is satisfied. Because we matched treated and control voters based on turnout prior to the ticket, this is unsurprising. The visual results of the post-treatment trend corroborate what we found nationwide using the municipal-level data; namely, ticketing has a different effect for Black voters than for others..

>>>>

---
title: "Policing for Profit and the Undermining of Democracy"

author:
- Jonathan Ben-Menachem^[PhD Student, Columbia University, Department of Sociology (jb4487@columbia.edu)]
- Kevin Morris^[PhD Student, CUNY Graduate Center, Department of Sociology (kmorris@gradcenter.cuny.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "policing_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  The American criminal legal system is an important site of political socialization: sociologists show how the alienating effect of criminalization pushes people away from public institutions, and political scientists have measured the effects of policing and incarceration on political participation. Despite this burgeoning literature, no research has directly investigated how ticketing for low-level offenses affects political participation. Using data from the Census of Governments and the national voter file, we test the causal effect of fines and fees practices on American voting behavior at the city level (aggregate). Additionally, we exploit individual-level ticketing data from Hillsborough County, Florida between 2003 and 2019 to determine whether getting a ticket impacts political participation. Nationally, we estimate that each 10 percent increase in city fines and fees collections between 2012 and 2017 is associated with a roughly 0.6 percentage point turnout decrease but was not associated with Black turnout. In Tampa, Florida, we estimate that being ticketed decreases an individual’s likelihood of voting in federal elections by 1.3 percentage points, but we do not uncover evidence of a distinct political effect for Black voter turnout. In local elections, however,   
  |
  |
  **Keywords**: voting, monetary sanctions, race, punishment, political participation
nocite: | 
  @Burch2011, @Burch2013, @White2019a

@@

Work exploring how criminalization directly and indirectly influences political participation has exploded in recent years. Scholars have found that criminal legal contact (i.e. arrest, conviction, incarceration) consistently discourages voting [@Weaver2010; @Burch2011; @White2019a]. This work, however, has largely focused on the effects of highly consequential contact with the criminal legal system such as incarceration and felony convictions [@Burch2014; @Lee2014]. Our project represents the first use of individual-level administrative data to identify the causal effect of ticketing on voter behavior.

The use of administrative data marks an important step forward in our understanding of how low-level contact with the criminal legal system structures participation. Past work looking at the individual-level effects of low-level contact has relied entirely on survey or qualitative data [e.g. @Walker2014; @Weaver2010]. Their in-depth research allows for the testing of specific psychological mechanisms and personal interpretations of criminal legal contact, but does not allow us to generalize more broadly. As Weaver and Lerman [-@Weaver2010, page 821] note, it may also introduce measurement bias. Our analysis investigates actual voting behavior following actual ticketing events, not reported voting behavior or reported exposure to police ticketing. The administrative data therefore allow us both to sidestep reporting error and to observe the behavior of the 232,568 individuals ticketed over a 6 year period---a far larger pool than even the most robust surveys.

We begin by joining fees and fines data from municipalities around the country in 2012 and 2017 with a novel dataset estimating racial turnout for each municipality in the country. While we find that increased fees and fines likely cause lower turnout among non-Black Americans, we find no effect of fees and fines on Black turnout.

Because our national analysis does not allow us to directly observe the causal effect of police ticketing on voter turnout, we use individual-level traffic stop data from Hillsborough County, Florida, to identify the turnout patterns of voters who were ticketed between the 2012 and 2018 elections. By matching individual voters who were ticketed to similar voters who were ticketed at later points and running a difference-in-differences model, we estimate the causal effect of these tickets on turnout. We find that being ticketed reduces the chance that an individual will turn out in the subsequent election.

We demonstrate that police ticketing practices---the most widespread form of police contact in America---influence the political behavior of Americans. This finding supports and extends Vesla Weaver and Amy Lerman’s theory of criminal legal contact discouraging political participation [@Weaver2010]. Our findings are relevant for interdisciplinary scholars of crime, race, politics, political economy, social movements, and policing.

@@

An emerging political science literature directly investigates the relationship between criminalization and political behavior. Additionally, sociologists have incorporated contemporary fines and fees practices into a broader empirical-historical theory of racialized punishment policy [@Harris2010; @Friedman2020].

@@

@@

Using existing national survey data and extended interviews, Vesla Weaver and Amy Lerman found that trust in government and willingness to vote decrease as individuals progress through increasingly intense levels of criminal legal contact (questioned by police, arrested, convicted, incarcerated) [@Weaver2010]. Weaver and Lerman also investigate the potential role of criminal legal contact in facilitating a collective political consciousness among Black Americans, or the concept of a “linked fate” where “one’s individual opportunities are linked to the entire racial group” [@Lerman2014, page 197]. Their interview data lead them to conclude that while criminal legal contact does increase perceptions of linked fate among Black respondents, “a custodial linked fate is not the building block of greater activism and a mobilized racial identity; instead, it is demobilizing”---and interview respondents who had been incarcerated reported that withdrawal from public institutions and democratic participation constituted a form of self-preservation [@Lerman2014].

@@

It may be the case that Black and non-Black Americans employ different self-preservation strategies, however. @Walker2020 makes clear that, when Black Americans locate their contact with the criminal legal system in a narrative of racial injustice, they can be politically mobilized; on the other hand, @Lerman2014 shows that Black individuals’ ability to situate their life experiences in a larger legacy of exploitation can be demobilizing. Turning to administrative data analysis, Ariel White’s analysis of how short jail sentences affect voting in Harris County, Texas found that Black defendant turnout in subsequent elections dropped by 13 percentage points---but the effect on white turnout was not statistically significant [@White2019].

@@

Walker found that nonwhite people who had experienced *either* proximal or personal contact *at any level of intensity*, including being stopped and questioned by police, were mobilized to take part in nonvoting political activity---and the same held true for whites, but with a much smaller effect size. Walker theorizes that this finding suggests that nonwhite people view criminalization “as a form of systemic political threat,” an analysis comparable to “linked fate.” Additionally, Walker’s survey questions excluded “minor traffic stops,” the primary focus of this paper.

@@

Although geographic or spatial analyses of how criminalization affects political socialization technically move beyond proximal contact as an extant conceptual framework, similar mechanisms should operate on the neighborhood level. Traci Burch found that North Carolina block groups with higher concentrations of people who had experienced incarceration or criminal legal supervision saw substantial reductions in voter turnout, and confirmed her findings at the individual level [@Burch2014]. @Morris2020 similarly finds that neighborhoods with more active voters lost to felony disenfranchisement turned out at lower rates in the 2017 mayoral race in NYC, and that this effect was particularly pronounced in Black neighborhoods.

@@

Over the past few years, scientific understanding of the political effects of more serious interactions with the criminal legal system has grown by leaps and bounds. The following table summarizes our understanding of the literature. As the table makes clear, a large body of research has now demonstrated the negative effects of high-level contact with the criminal legal system on voter turnout. Although lower-level stops are far more frequent, our knowledge of their (de)mobilizing effects is under-developed.

@@

We argue that fines and fees invoke a distinct historical legacy of racist exploitation that merits further study.

@@

Racial capitalist political economy imposes fiscal constraints regardless of city officials’ intentions. In their study of fines and fees reliance in Chicago suburbs, Josh Pacewicz and John Robinson, III show how the racialization of municipal opportunity structures prevents majority-Black suburban governments from acquiring sufficient revenue from “good” sources such as sales taxes or property taxes, catalyzing a reliance on fines and fees to fund government services [@Pacewicz2020]. Pacewicz and Robinson III note that these racialized municipal opportunity structures are tied to the political-historical processes of neoliberalism and fiscal austerity, an observation that aligns with recent scholarship detailing the origins of fees paid by incarcerated people [@Kirk2020]. The late 20th century “property tax revolt” most significantly benefited white homeowners [@Martin2017], and without property tax hikes as a revenue-raising option, smaller cities and towns have few choices.

@@

With this history in mind, we argue that Black Americans can situate exploitative ticketing practices in a longer historical legacy of collective disparate treatment and violence at the hands of the state, which may facilitate a stronger collective political analysis when police ticketing practices become more extreme. The historical use of fees and fines make this sort of police contact especially likely to evoke such narratives. Accordingly, we expect that Black voters’ political behavior in response to police ticketing will be distinct from the political behavior of white voters.

@@

**Census of Governments**: Every five years, the US Census Bureau collects data from governmental bodies around the country. We use two snapshots of the COG: one that covers the years 2008 -- 2012, and another that covers 2013 -- 2017. The COG asks municipalities to report how much revenue they collect from “penalties imposed for violation of law; civil penalties (e.g., for violating court orders); court fees if levied upon conviction of a crime or violation... and forfeits of deposits held for performance guarantees or against loss or damage (such as forfeited bail and collateral).” We divide these amounts by the population to estimate the per-capita fees and fines revenue raised in each municipality. This data has been used as a proxy for the amount of tickets issued at the local level in scholarly research [see, for instance, @Goldstein2020; @Sances2017].

The COG data are useful to isolate ticketing practices because other fines and fees (e.g. fines imposed at sentencing in state court, probation fees, conviction fees, public defender fees) are typically collected by county or state governments. The COG data only include local revenues, which means the share of fines revenue corresponds more directly to local police practices. Following the guidance of @Sances2017, we retain only municipalities with police and/or court systems, as these are the only municipalities capable of issuing tickets. We also only retain municipalities with populations of at least 2,500.

**Registered Voter Files**: The COG data do not include estimates of turnout at the local level. To estimate local turnout, we leverage registered voter files provided by data vendor L2 Political. In Washington, DC, and every state in the country (with the exception of North Dakota) voters are required to register to vote prior to participating. These files are considered public record, and in most states contain information about each voter’s age, residential address, political affiliation, and participation history. L2 collects these voter files from around the country and merges them with proprietary data that estimates other voter- and household-level characteristics such as race and education. L2 also geocodes voters’ residential addresses. L2 does not collect all states’ voter file on the same date; the date of the “raw” voter file from each state from which the L2 data are modelled can be found in Appendix A.

We merge these geocoded voter files with “place” and “county subdivision” shapefiles published by the US Census Bureau. By aggregating these individual-level records to the municipal level, we are able to estimate the number of registered voters and the number of ballots cast in every city in the country. Moreover, by using the racial estimates provided by L2 we can estimate the number of ballots cast by voters of each race in each city. Each voters’ race is modeled based on L2’s algorithms and not self-reported, surely leading to errors at the individual-voter level. We assume this does not present a problem when records are aggregated to the municipal level.

We calculate turnout by dividing the number of ballots cast according to L2 by the 5-year estimate of each municipality’s citizen voting-age population (CVAP) from the US Census Bureau. These estimates are available for individual racial groups. Because our turnout numerator and denominators come from different sources there is slight measurement error: our estimate indicates, for instance, that roughly 1 percent of municipalities had Black turnout above 100 percent. Where more votes were cast than the Census estimate for CVAP, we recode turnout as 100 percent. Our results do not change if we instead simply remove these municipalities, or use unmodified turnout estimates.

**Estimation Strategy**: We begin by merging each municipality’s fees and fines per capita (that is, dollars per resident) from the 2012 COG to its turnout from the 2014 midterm election, and the 2017 COG to its 2018 turnout. Because this measure is highly right-skewed, we use the natural log of fees and fines per capita plus one. We also control for other municipal and state characteristics. These include racial composition, population density, median income, collegiate education, median age, and share of the population over 64 years of age from the 5-year American Community Survey ending in 2014 and 2018, and total municipal revenue, share of revenue that comes from taxes, and share of each municipality’s revenue that comes from the state or federal government in the 2012 and 2017 COGs. We retain the 8,954 municipalities whose data is reported to the COG in both 2012 and 2017 and meet the other criteria noted above.

The 5-year cycle of the COG does not neatly line up with the general 4-year cycle typical of American elections. The timing of the COG and midterm elections mean that our analytical strategy is conservative. If municipalities shift their fees and fines collections dramatically over the 2013 -- 2017 period, this shift may influence their 2014 turnout. However, we assume that turnout in 2014 will still be influenced by the policing practices of the 2008 -- 2012 period, while 2018 turnout will be largely influenced by the practices of the 2013 -- 2017 period.

We  run a two-way fixed effects model to explore whether changes in fees and fines collections from 2012 to 2017 were associated with changes in turnout, and whether these effects differed for Black and non-Black voters.

@@

**Hypothesis 1:** Increased reliance on fines and fees revenue at the city level reduces voter turnout at the city level.*

**Hypothesis 2:** The effect of the city-level relationship between fines and voter turnout is stronger for Black voters, given the historical and explicit racial injustice narratives used to explain these practices.*

@@

Our national test of the relationship between municipal fees and fines collection and turnout offers novel insight into the political socialization of low-level police contact. It cannot, however, directly observe the causal mechanism we argue is at play: namely, that direct contact with the criminal legal system influences turnout. If interpreted in isolation, our national analysis would suffer from the ecological fallacy. In other words, it is possible that any observed relationship between fees and fines collections and municipal turnout is being driven by the behavior of voters other than those who are ticketed. Moreover, there are many other factors that could be at play. Increased fees and fines could be correlated with increased policing in general, for instance. We are also unable to isolate the individual-level effect of being ticketed on turnout using this setup.

Another drawback of the national analysis is that it can't distinguish personal contact from proximal contact (“having a loved one who is a custodial citizen without yourself having had contact”) [@Walker2017]. Because existing research investigating the effect of proximal contact has produced mixed results [@Burch2014; @Lee2014; @Sugie2015; @White2019; @Walker2017; @Walker2020], we are unable to prove that personal police contact is the only dynamic driving effects observed in our national analysis. For that reason, we turn to individual-level data from Hillsborough County, Florida (home to Tampa).

Hillsborough County is interesting for theoretical reasons: Tampa is in the top decile of municipalities in terms of fees and fines increase between the 2012 and 2017 COGs, and its population puts it in the top 1 percent of cities nationwide. It also has a relatively large Black population, allowing us to test for racial differences in the turnout effect of ticketing.

@@

Hillsborough County also makes a unique set of data publicly available. The county clerk makes information available about every traffic stop in the county going back to 2003. This data includes the name and date of birth of the individual stopped; the date of the offense; and the amount the individual paid for a ticket.

To identify ticketed voters, we match the first and last names and dates of birth from these data against the Hillsborough County registered voter file. Using a single post-treatment snapshot of the voter file can result in conditioning on a post-treatment status [see @Nyhan2017]. Instead, we collect snapshots of the voter file following each even-year general election between 2012 and 2018.^[Because such frequent snapshots are not available from L2, we here use the registered voter file made freely available by the state of Florida.] We thus observe virtually all individuals who were registered to vote at any time during our period of study. Unique voter identification numbers allow us to avoid double-counting voters who are registered in multiple snapshots. We retain each voter’s earliest record, and geocode voters to their home census block groups. Because we argue that police contact is the driving force behind any turnout effect, we remove tickets issued by red-light cameras.

@@

By matching the police stop and voter records, we identify all voters who were ticketed between the 2012 and 2018 general elections. Voters are considered “treated” in the general election following their ticket.

These voters are then matched to a control voter using a nearest-neighbor approach, with a genetic algorithm used to determine the best weight for each characteristic [@Sekhon2011]. As we show in Appendix XX, there is real reason to believe that individuals who are ticketed differ in meaningful ways for which we cannot directly control. As such, control voters are individuals who are stopped in the two years following the post-treatment election of the treated voters. Put differently, if a voter is stopped between 2012 and 2014, their control voter must be an individual stopped between the 2014 and 2016 elections. A voter cannot both be a treated and control voter for the same election; therefore, someone ticketed between the 2012 and 2014 elections and again between the 2014 and 2016 elections cannot serve as a control for anyone ticketed between 2012 and 2014. We assume that after controlling for observable characteristics, psat turnout, *and* the unobservable characteristics associated with receiving a ticket the timing of the ticket is effectively random. In short, any turnout difference between the treated voters and their controls is the causal effect of a police stop on turnout.

We match voters on individual-level characteristics (race / ethnicity; gender; party affiliation; age; registration date; number of traffic stops between 2003 and the election prior to the treatment period) and block group-level characteristics from the 2012 5-year ACS estimates (median income; share of the population with some college; unemployment rate). We also match exactly on whether an individual was ticketed for a civil or criminal infraction, as well as whether they eventually paid a fine. We also match treated and control voters on their turnout in the three pre-treatment elections.

Our analytical approach thus incorporates two different causal approaches. Leveraging pre-treatment turnout allows us to estimate a traditional difference-in-differences model. Difference-in-difference models assume that, in the absence of the treatment, any turnout gap between treated and control voters would have remained stable in the post-treatment period. While the validity of this assumption is commonly established by examining pre-treatment trends in turnout between the two groups, it can be improved upon by using the matching methodology described above. In the case at hand, we both establish that the turnout gap between treatment and control voters was constant prior to treatment, *and* that the observable characteristics of treated and control voters are similar---thus lending credence to the assumption that this gap would have remained constant absent treatment. Finally, by drawing controls from the population of individuals who were ticketed in subsequent years, we also indirectly control for any unobserved characteristics associated with receiving a ticket. For a more detailed discussion of how matching can improve upon traditional difference-in-difference approaches when using panel data, see @Imai2020.

It is important to note that while our primary analyses include all individuals who were stopped, our results hold even when we shrink the window around an election dramatically. As we show in Appendix XX, voters ticketed in the two months before an election turned out at lower rates than matched controls ticketed in the two months following an election.

We then estimate the following equation:

\begin{gather}
\label{eq:1}
v_{it}=\beta_0+\beta_1Treated_{i}+\beta_2Post Treatment_{t} + \nonumber \\
\beta_3Treated_{i}\times Post Treatment_{t} + \\
\delta{Z}_{i} + \mathcal{E}_{it}. \nonumber
\end{gather}

Individual *i*'s turnout (*v*) in year *t* is a function of the year and whether they were stopped by the police. In the equation, *$\beta$~1~Treated~i~* measures the historical difference between treated voters and their controls. *$\beta$~2~Post Treatment~t~* measures whether turnout increased for controls in the first election following the treated voter’s stop, while *$\beta$~3~Treated~i~ × Post Treatment~t~* tests whether turnout changed differently for treated voters than their controls in the election following their police stop. *$\beta$~3*, then, will capture the causal effect of a police stop on voter turnout. Finally, the matrix *$\delta$Z~i~* contains the individual- and neighborhood-level characteristics on which the match was performed, included in some of the models. Equation 1 allows us to test Hypotheses 3:

**Hypothesis 3**: Being ticketed reduces an individual’s likelihood of turning out in the subsequent election.

We also estimate Equation 2, in which the treatment effect is interacted with whether a voter is Black (*$\beta$~4~Treated~i~ × Post Treatment~t~ × Black~i~*):

\begin{gather}
\label{eq:2}
v_{it}=\beta_0+\beta_1Treated_{i}+\beta_2Post Treatment_{t} + \nonumber \\
\beta_3Treated_{i}\times Post Treatment_{t} + \beta_4Treated_{i}\times Post Treatment_{t}\times Black_{i} +\\
\delta{Y}_{i} + \delta{Z}_{i} + \mathcal{E}_{it}. \nonumber
\end{gather}

where *$\delta$Y~it~* includes the dummy for Black interacted with the treatment and post treatment dummies.

Equation 2 allows us to test Hypotheses 4:

**Hypothesis 4**: The effect of being ticketed on an individual's turnout is less negative / more positive if that individual is Black.

Because Tampa holds off-year local elections, we are able to test whether ticketed voters target officials who are able to change local police practices. @Laniyonu2019 gives us reason to believe that the effect should be stronger for local elections.

**Hypothesis 5**: Any potential mobilizing turnout effect in Tampa is stronger for local (mayoral) elections, since those elections directly affect police practices.

@@

Figure \ref{fig:did1} makes a number of things immediately apparent. First, we can see that for both Black and non-Black voters the parallel trends assumption is satisfied. Because we matched treated and control voters based on turnout prior to the ticket, this is unsurprising.
